Iterative Monomorphisation

### 2025-02-13 iterative_monomorphisation.tex
contrived
rename

\author[1,2]{Tanguy Bozec}[%
email=tanguy.bozec@ens-paris-saclay.fr,
]
\author[2]{Jasmin Blanchette}[%
email=jasmin.blanchette@ifi.lmu.de,
%url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
]
\address[1]{ENS Paris-Saclay, Université Paris-Saclay, France}
\address[2]{Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}

\begin{abstract}
Monomorphisation can be used to extend monomorphic provers to support polymorphic logics. We propose an iterative approach, which is necessarily incomplete but which works well in practice. It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation demonstrates that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.
\end{abstract}

\begin{keywords}
   Polymorphism\sep
   monomorphism\sep
   automatic theorem proving
\end{keywords}

@@

This example illustrates how an infinite number of new formulae can be generated from a simple initial problem. Although the example may seem artificial, similar problems frequently arise in practice. For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior. Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae to be useful.

@@

\begin{table}[ht]
\centering\begin{tabular}{@{}lccc@{}}
   \toprule
   & Native & Mono & Union \\
   \midrule
   E              & -- & 340 & 340 \\
   Leo-III with E &  157 & 231 & 274 \\ 
   Zipperposition & 339 & 351 & 404 \\%[1.5\jot]
%   Total          & 0 & 0 & 0 \\
   % Satallax       &  & 0 & 0 \\ % doesn't understand negated conjectures
   %Leo-III &  68 with cvc4 & 0 & 0 \\ % same as without anything for the native case
   %Vampire & 0 & 0 & 0 \\ % Vampire's parser fails on too many polymorphic problems (about 1/3 of them)
   \bottomrule
   \label{native_vs_mono_eval}
\end{tabular}
\caption{Evaluation of native polymorphism vs.\ monomorphisation}
\end{table}

@@

% target page number: 0.5-1
%\section{Related work}
%\label{sec:related-work}
%
%  * monomorphisation:
%    Isabelle for \textit{smt} tactic (Böhme), \textit{metis}, Sledgehammer
%  * a precursor for this work is the \verb|POLY_ASSUME_TAC| preprocessor of
%    MESON for HOL Light. It is unfortunately not documented beyond the comment
%    ``Push duplicated copies of poly theorems to match existing assumptions''
%    but it appears to perform at least partial monomorphisation.
%
%  * alternative: encode polymorphism
%    * many approaches to encode monomorphic and polymorphic types, starting
%      with Enderton \cite{xxx}, Stickel \cite{xxx}, and Wick and McCune
%      \cite{xxx}
%    * Blanchette et al.~\cite{xxx} exploit ``monotonicity'' to erase type
%      information
%    * see Blanchette et al.~\cite[Section~9]{xxx} for a more detailed coverage
%      of encodings of polymorphism
%    
%%\li{https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette/enc_types_article.pdf}

@@

We see the following avenues for future work. First, iterative monomorphisation
blindly enumerates candidate instantiations, without exploiting any knowledge
about the logical structure of the formulas in which symbols occur. For
example, a lemma $\sym{p}\langle\alpha\rangle$ cannot be used to prove the
conjecture $\lnot \sym{p}\langle\ty{nat}\rangle$ because of the incompatible
polarities, but our procedure instantiates $\alpha$ with $\ty{nat}$ regardless.
Second, some automatic provers as well as tools like Sledgehammer include a
relevance filter that heuristically selects a subset of the available axioms;
filters such as MePo \cite{meng-paulson-2009} and SInE
\cite{hoder-voronkov-2011} are iterative and could be interleaved with
monomorphisation. Third, although one would expect native implementations of
polymorphism to outperform any preprocessor, our evaluation clearly shows that
this is not the case, therefore indicating that there is considerable room for
improvement on the native front.

\section*{Acknowledgements}

@@

This research is co-funded by the European Union (ERC, Nekoka, 101083038).
Views and opinions expressed are however those of the authors only and do not
necessarily reflect those of the European Union or the European Research
Council. Neither the European Union nor the granting authority can be held
responsible for them.

\bibliography{citations}

>>>>

\author{Tanguy Bozec\inst{1,2} \and
Jasmin Blanchette\inst{2}\orcidID{0000-0002-8367-0936}}

\institute{ENS Paris-Saclay, Université Paris-Saclay, France \and
Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}

@@

This example illustrates how an infinite number of new formulae can be generated from a simple initial problem. Although the example may seem contrived, similar problems frequently arise in practice. For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior. Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae to be useful.

@@

\begin{table}[ht]
\caption{Evaluation of native polymorphism vs.\ monomorphisation}
\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & Native & Mono & Union \\
   \midrule
   E              & -- & 340 & 340 \\
   Leo-III with E &  157 & 231 & 274 \\ 
   Zipperposition & 339 & 351 & 404 \\%[1.5\jot]
%   Total          & 0 & 0 & 0 \\
   % Satallax       &  & 0 & 0 \\ % doesn't understand negated conjectures
   %Leo-III &  68 with cvc4 & 0 & 0 \\ % same as without anything for the native case
   %Vampire & 0 & 0 & 0 \\ % Vampire's parser fails on too many polymorphic problems (about 1/3 of them)
   \bottomrule
   \label{native_vs_mono_eval}
\end{tabular}
\end{table}

@@

% target page number: 0.5-1
%\section{Related work}
%\label{sec:related-work}
%
%  * monomorphisation:
%    Isabelle for \textit{smt} tactic (Böhme), \textit{metis}, Sledgehammer
%  * a precursor for this work is the \verb|POLY_ASSUME_TAC| preprocessor of
%    MESON for HOL Light. It is unfortunately not documented beyond the comment
%    ``Push duplicated copies of poly theorems to match existing assumptions''
%    but it appears to perform at least partial monomorphisation.
%
%  * alternative: encode polymorphism
%    * many approaches to encode monomorphic and polymorphic types, starting
%      with Enderton~\cite{xxx}, Stickel~\cite{xxx}, and Wick and McCune
%     ~\cite{xxx}
%    * Blanchette et al.~~\cite{xxx} exploit ``monotonicity'' to erase type
%      information
%    * see Blanchette et al.~~\cite[Section~9]{xxx} for a more detailed coverage
%      of encodings of polymorphism
%    
%%\li{https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette/enc_types_article.pdf}

@@

We see the following avenues for future work. First, iterative monomorphisation
blindly enumerates candidate instantiations, without exploiting any knowledge
about the logical structure of the formulas in which symbols occur. For
example, a lemma $\sym{p}\langle\alpha\rangle$ cannot be used to prove the
conjecture $\lnot \sym{p}\langle\ty{nat}\rangle$ because of the incompatible
polarities, but our procedure instantiates $\alpha$ with $\ty{nat}$ regardless.
Second, some automatic provers as well as tools like Sledgehammer include a
relevance filter that heuristically selects a subset of the available axioms;
filters such as MePo~\cite{meng-paulson-2009} and SInE
~\cite{hoder-voronkov-2011} are iterative and could be interleaved with
monomorphisation. Third, although one would expect native implementations of
polymorphism to outperform any preprocessor, our evaluation clearly shows that
this is not the case, therefore indicating that there is considerable room for
improvement on the native front.

\begin{credits}
   \subsubsection{\ackname}

@@

   This research is co-funded by the European Union (ERC, Nekoka, 101083038). Views and opinions expressed are however those of the authors only and do not necessarily reflect those of the European Union or the European Research Council. Neither the European Union nor the granting authority can be held responsible for them.
\end{credits}

\bibliographystyle{splncs04}
\bibliography{citations}
